Vagueness \& Rationality

### 2018-08-19 springer/current/e1.tex
finished renumbering

%However, the picture painted by these findings might be misleading. First of all, there is a possibly confounding factor. Contrast, from Experiment 1, an expression from the vague condition: `the square with few dots' with an expression from the crisp condition: `the square with 5 dots'. One difference is that `few' has the potential for vagueness, whereas `5' is crisp. But another difference is that `few' is verbal while `5' is numerical, in the sense that a number is mentioned explicitly. Since these two differences could not be separated in Experiment 1, the vagueness advantage finding is vulnerable to an alternative interpretation, that what we saw as a vagueness advantage was in contrast an advantage for the verbal form of the quantifier. In Experiment 2 we therefore created verbal and numeric versions of each of the vague and crisp instructions so that we could compare vague and crisp conditions while taking account of verbal / numeric format.

%Another potential problem with Experiment 1 is the following. Participants chose one of two squares: therefore the `vague' quantifiers (e.g., `few') uniquely identified one square. Recall our definition of vague  -- ``a word is precise if it describes a well-defined set of objects. By contrast, a word is vague if it is not precise''.  In Experiment 1, the quantifiers in the vague conditions did not realise their potential for vagueness. This is because there were no borderline cases of the referent that could make the referent set `not well-defined', and perhaps because using definite articles in the instructions implied that only one option was correct. Using error feedback in Experiment 1 could have exacerbated this.

\marginnote{C\_Exp\_1}

To find out what happens when words are used in a context where their potential for vagueness comes to the fore, Experiment 2 used three arrays (rather than two) so that the vague description had more than one possible referent, and used indefinite articles to avoid the impression that only one response counted as correct, and was carried out without error feedback.
An indication that the potential for vagueness was realised in Experiment 2 is that the borderline response was chosen fairly often: 16\% of the time.

In Experiment 2, an item was a referring expression instruction followed by a set of three dot arrays defined by a triple of numbers, representing the number of dots in the left, middle, and right arrays. We used four different triples of numbers: (6,15,24); (16,25,34); (26,35,44); (36,45,54). Each set of arrays comprised three arrays (instead of two as in Experiment 1); the array representing the central number was always presented in the middle of the three; there were two flanking arrays where one had fewer dots than the central array and the other had more.

@@

\begin{table}
\centering
\caption{Experiment 1 instructions arranged by condition. The instructions given in the table started with ``Choose \ldots"}
\label{instructionse2}
\begin{tabular}{cccll}
\hline\noalign{\smallskip}
Item & Quantity & Number & Crisp & Vague\\
\noalign{\smallskip}\hline\noalign{\smallskip}
06:15:24 & Small & Numeric & the square with 6 dots & a square with about 10 dots\\
06:15:24 & Small & Verbal & the square with the fewest dots & a square with few dots\\
06:15:24 & Large & Numeric & the square with 24 dots & a square with about 20 dots\\
06:15:24 & Large & Verbal & the square with the most dots & a square with many dots\\
\noalign{\smallskip}\hline\noalign{\smallskip}
16:25:34 & Small & Numeric & the square with 16 dots & a square with about 20 dots\\
16:25:34 & Small & Verbal & the square with the fewest dots & a square with few dots\\
16:25:34 & Large & Numeric & the square with 34 dots & a square with about 30 dots\\
16:25:34 & Large & Verbal & the square with the most dots & a square with many dots\\
\noalign{\smallskip}\hline\noalign{\smallskip}
26:35:44 & Small & Numeric & the square with 26 dots & a square with about 30 dots\\
26:35:44 & Small & Verbal & the square with the fewest dots & a square with few dots\\
26:35:44 & Large & Numeric & the square with 44 dots & a square with about 40 dots\\
26:35:44 & Large & Verbal & the square with the most dots & a square with many dots\\
\noalign{\smallskip}\hline\noalign{\smallskip}
36:45:54 & Small & Numeric & the square with 36 dots & a square with about 40 dots\\
36:45:54 & Small & Verbal & the square with the fewest dots & a square with few dots\\
36:45:54 & Large & Numeric & the square with 54 dots & a square with about 50 dots\\
36:45:54 & Large & Verbal & the square with the most dots & a square with many dots\\
\noalign{\smallskip}\hline
\end{tabular}
\end{table}

@@

We formulated the following hypotheses for Experiment 2:

@@

Means for response times and proportion of borderline responses are given in Fig. (\ref{resultse2}). 
Response times from all trials were trimmed at 2.5 standard deviations for each subject, leading to the loss of 236 trials, 3.1\% of the data. 
A linear mixed model was constructed for the (logged) response times, 
with sum-coded vagueness, instruction format, (and their interaction), and item as fixed effects, and the same effects as slopes over participant for random effects. We found:

@@

A generalized linear mixed model \citet{jaeger2008categorical} was fit to the data for selection of the borderline response, with sum-coded vagueness, instruction format, (and their interaction), and item as fixed effects, and the same effects as slopes over participant for random effects. The distribution of responses over the nearest match square, the borderline square, and the furthest match square are given in Fig, \ref{resultse2}. Participants chose the borderline square on $16.6\%$ of trials overall.

@@

\begin{figure}[htbp]
\centering
\includegraphics[trim = 20mm 0mm 35mm 0mm, clip, width=.49\textwidth]{figures/e2-rtplot-1.pdf}
\includegraphics[trim = 20mm 0mm 35mm 0mm, clip, width=.49\textwidth]{figures/e2-blBarChart-1}
\caption{Experiment 2 results: mean response times by condition and item, and counts of borderline case responses by condition.}
\label{resultse2}
\end{figure}

@@

Experiment 2 tested to see whether when borderline cases are present, vague instructions would speed responses as they did in Experiment 1 when there were no borderline squares. 
We actually found a small (but statistically significant) \emph{disadvantage} of vague instructions: vague instructions slowed people down by 112 ms on average. We also found that the effect of instruction format was significant, with numerical format slowing responses by 689 ms on average, such that the disadvantage of numerical format overwhelmed the contribution of vagueness. The \emph{verbal vague} condition was still responded to faster than the \emph{numerical crisp} condition, so the pattern from Experiment 1 was reproduced, but in the light of the evidence from Experiment 2, in the presence of borderline cases, the advantage that was ascribed to vagueness before now looks more like an advantage of verbal instruction format.

However, once again there is a possibly confounding factor. Observe that, in Experiment 2, instruction format (i.e., the choice between numeric and verbal) went hand in hand with might be called the (human) {\bf selection algorithm}: To see this, consider the task of selecting the dot array that contains ``few dots'': to do this, it suffices to {\em compare} the three arrays and select the one that contains the fewest elements.  To select the dot array that contains ``16 dots'' seems to require the participant to estimate, and then {\em match}, the cardinality of (at least) one dot array to 16, a process which could plausibly take longer, independently of vagueness. Therefore, our results so far permit the interpretation that what made the instructions in the verbal condition fast is not the fact that they were worded verbally, but that they allowed participants to use a comparison ``algorithm''.

In the next two experiments we pitted the comparison algorithm and matching algorithm selection tasks against each other while controlling vagueness and instruction format. In Experiment 3 we restricted all the instructions to \emph{numeric} quantifiers while factorially manipulating vagueness and selection task. In Experiment 4 we ensured that all instructions used \emph{verbal} quantifiers, while also factorially manipulating vagueness and selection task. This allowed us to distinguish between the predictions of the selection task account and the instruction format account.

>>>>

\marginnote{C-Exp-1}

To find out what happens when words are used in a context where their potential for vagueness comes to the fore, Experiment 1 used three arrays (rather than two) so that the vague description had more than one possible referent, and used indefinite articles to avoid the impression that only one response counted as correct, and was carried out without error feedback. An indication that the potential for vagueness was indeed realised in Experiment 1 is that the borderline response was chosen fairly often: 16\% of the time.

In Experiment 1, an item was a referring expression instruction followed by a set of three dot arrays defined by a triple of numbers, representing the number of dots in the left, middle, and right arrays. We used four different triples of numbers: (6,15,24); (16,25,34); (26,35,44); (36,45,54). Each set of arrays comprised three arrays (instead of two as in Experiment 1); the array representing the central number was always presented in the middle of the three; there were two flanking arrays where one had fewer dots than the central array and the other had more. An example stimulus is given in Figure \ref{Experiment1and2examplestimulus}.

@@

\begin{table}
\centering
\caption{Experiment 1 instructions arranged by condition. The instructions given in the table started with ``Choose \ldots"}
\label{instructionsC-exp-1}
\begin{tabular}{cccll}
\hline\noalign{\smallskip}
Item & Quantity & Number & Crisp & Vague\\
\noalign{\smallskip}\hline\noalign{\smallskip}
06:15:24 & Small & Numeric & the square with 6 dots & a square with about 10 dots\\
06:15:24 & Small & Verbal & the square with the fewest dots & a square with few dots\\
06:15:24 & Large & Numeric & the square with 24 dots & a square with about 20 dots\\
06:15:24 & Large & Verbal & the square with the most dots & a square with many dots\\
\noalign{\smallskip}\hline\noalign{\smallskip}
16:25:34 & Small & Numeric & the square with 16 dots & a square with about 20 dots\\
16:25:34 & Small & Verbal & the square with the fewest dots & a square with few dots\\
16:25:34 & Large & Numeric & the square with 34 dots & a square with about 30 dots\\
16:25:34 & Large & Verbal & the square with the most dots & a square with many dots\\
\noalign{\smallskip}\hline\noalign{\smallskip}
26:35:44 & Small & Numeric & the square with 26 dots & a square with about 30 dots\\
26:35:44 & Small & Verbal & the square with the fewest dots & a square with few dots\\
26:35:44 & Large & Numeric & the square with 44 dots & a square with about 40 dots\\
26:35:44 & Large & Verbal & the square with the most dots & a square with many dots\\
\noalign{\smallskip}\hline\noalign{\smallskip}
36:45:54 & Small & Numeric & the square with 36 dots & a square with about 40 dots\\
36:45:54 & Small & Verbal & the square with the fewest dots & a square with few dots\\
36:45:54 & Large & Numeric & the square with 54 dots & a square with about 50 dots\\
36:45:54 & Large & Verbal & the square with the most dots & a square with many dots\\
\noalign{\smallskip}\hline
\end{tabular}
\end{table}

\begin{figure}[htbp]
\centering
\includegraphics[width=.75\textwidth]{figures/Ce1-example-screenshot}
\caption{Experiment 1 and 2 example stimulus}
\label{Experiment1and2examplestimulus}
\end{figure}

@@

We formulated the following hypotheses for Experiment 1:

@@

Means for response times and proportion of borderline responses are given in Fig. (\ref{resultsC-exp-1}). Response times from all trials were trimmed at 2.5 standard deviations for each subject, leading to the loss of 236 trials, 3.1\% of the data. A linear mixed model was constructed for the (logged) response times, with sum-coded vagueness, instruction format, (and their interaction), and item as fixed effects, and the same effects as slopes over participant for random effects. We found:

@@

A generalized linear mixed model \citet{jaeger2008categorical} was fit to the data for selection of the borderline response, with sum-coded vagueness, instruction format, (and their interaction), and item as fixed effects, and the same effects as slopes over participant for random effects. The distribution of responses over the nearest match square, the borderline square, and the furthest match square are given in Fig, \ref{resultsC-exp-1}. Participants chose the borderline square on $16.6\%$ of trials overall.

@@

\begin{figure}[htbp]
\centering
\includegraphics[trim = 20mm 0mm 35mm 0mm, clip, width=.49\textwidth]{figures/Ce1-rtplot-1.pdf}
\includegraphics[trim = 20mm 0mm 35mm 0mm, clip, width=.49\textwidth]{figures/Ce1-blBarChart-1}
\caption{Experiment 1 results: mean response times by condition and item, and counts of borderline case responses by condition.}
\label{resultsC-exp-1}
\end{figure}

@@

Experiment 1 tested to see whether when borderline cases are present, vague instructions would speed responses as they did in Experiment pilot exp 2 when there were no borderline squares.

We actually found a small (but statistically significant) \emph{disadvantage} of vague instructions: vague instructions slowed people down by 112 ms on average. We also found that the effect of instruction format was significant, with numerical format slowing responses by 689 ms on average, such that the disadvantage of numerical format overwhelmed the contribution of vagueness. The \emph{verbal vague} condition was still responded to faster than the \emph{numerical crisp} condition, so the pattern from Experiment pilot exp 2 was reproduced, but in the light of the evidence from Experiment 1, in the presence of borderline cases, the advantage that was ascribed to vagueness before now looks more like an advantage of verbal instruction format.

However, once again there is a possibly confounding factor. Observe that, in Experiment 1, instruction format (i.e., the choice between numeric and verbal) went hand in hand with might be called the (human) {\bf selection algorithm}: To see this, consider the task of selecting the dot array that contains ``few dots'': to do this, it suffices to {\em compare} the three arrays and select the one that contains the fewest elements.  To select the dot array that contains ``16 dots'' seems to require the participant to estimate, and then {\em match}, the cardinality of (at least) one dot array to 16, a process which could plausibly take longer, independently of vagueness. Therefore, our results so far permit the interpretation that what made the instructions in the verbal condition fast is not the fact that they were worded verbally, but that they allowed participants to use a comparison ``algorithm''.

In the next two experiments we pitted the comparison algorithm and matching algorithm selection tasks against each other while controlling vagueness and instruction format. In Experiment 2 we restricted all the instructions to \emph{numeric} quantifiers while factorially manipulating vagueness and selection task. In Experiment 3 we ensured that all instructions used \emph{verbal} quantifiers, while also factorially manipulating vagueness and selection task. This allowed us to distinguish between the predictions of the selection task account and the instruction format account.

### 2018-08-19 springer/current/e2.tex
finished renumbering

\marginnote{D\_Exp\_2}

The main aim of Experiment 2 was to see whether vagueness would exert beneficial effects when all conditions used numerals in the instructions, and when there were vague and crisp versions of the instructions for both comparison and matching strategies. The main changes from Experiment 2 were that the human selection task was explicitly controlled (i.e., in whether it amounted to matching or comparison), and that all conditions were constrained to mention a number. We used the same arrays as in Experiment 2. Table \ref{Instructions for e2} shows the instructions for each condition. Note the difference between ``fewer than 20'' and ``far fewer than 20'': whereas the former cannot have borderline cases (i.e., for each number it is clear whether the number is smaller than 20 or not), the latter can.

@@

\subsection{Hypotheses} % D_Exp_2
\noindent For Experiment 2, we hypothesised:

@@

\subsection{Experimental Method} % D_Exp_2

38 participants were recruited. We used a 2 x 2 factorial manipulation of vagueness and selection task (see Table \ref{Instructions for e2}).
On each trial an instruction was presented: participants pressed a key to dismiss the instruction, when the dot arrays were presented until the participant responded, and the response time and choice were recorded.

\subsection{Results} % D_Exp_2

@@

Response times were trimmed at 2.5 SD separately for each subject, leading to the loss of 204 trials (2.8\% of the trials).
Condition means for the remaining (logged) RTs are plotted in Figure \ref{resultse3}.
A linear mixed model was constructed for the logged response times, 
with sum-coded vagueness, instruction format, (and their interaction), and item as fixed effects, and the same effects as slopes over participant for random effects.

\begin{figure}[htbp]
\centering
\includegraphics[width=\textwidth]{figures/e3-rtplot-1.pdf}
\caption{Mean response times by condition for Experiment 2 where all instructions were numeric}
\label{resultse3}
\end{figure}

>>>>

\marginnote{D-Exp-2}

The main aim of Experiment 2 was to see whether vagueness would exert beneficial effects when all conditions used numerals in the instructions, and when there were vague and crisp versions of the instructions for both comparison and matching strategies. The main changes from Experiment 1 were that the human selection task was explicitly controlled (i.e., in whether it amounted to matching or comparison), and that all conditions were constrained to mention a number. We used the same arrays as in Experiment 1 (an example stimulus is given in Figure \ref{Experiment1and2examplestimulus}). Table \ref{Instructions for e2} shows the instructions for each condition. Note the difference between ``fewer than 20'' and ``far fewer than 20'': whereas the former cannot have borderline cases (i.e., for each number it is clear whether the number is smaller than 20 or not), the latter can.

@@

\subsection{Hypotheses}

For Experiment 2, we hypothesised:

@@

\subsection{Experimental Method}

38 participants were recruited. We used a 2 x 2 factorial manipulation of vagueness and selection task (see Table \ref{Instructions for e2}). On each trial an instruction was presented: participants pressed a key to dismiss the instruction, at which time the dot arrays were presented until the participant responded, and the response time and choice were recorded.

\subsection{Results}

@@

Response times were trimmed at 2.5 SD separately for each subject, leading to the loss of 204 trials (2.8\% of the trials). Condition means for the remaining (logged) RTs are plotted in Figure \ref{resultsD-exp-2}. A linear mixed model was constructed for the logged response times, with sum-coded vagueness, instruction format, (and their interaction), and item as fixed effects, and the same effects as slopes over participant for random effects.

\begin{figure}[htbp]
\centering
\includegraphics[width=\textwidth]{figures/De2-rtplot-1.pdf}
\caption{Mean response times by condition for Experiment 2 where all instructions were numeric}
\label{resultsD-exp-2}
\end{figure}

### 2018-08-19 springer/current/e3.tex
added a full table of instructiosn for D-Exp-2
finished renumbering

This experiment mirrors Experiment 3, but focussing on instructions that did not use a number. 
We manipulated vagueness and the selection task (comparison and matching). 
In order to implement the experiment without mentioning numbers in the instructions, we changed the sequence of each trial to include a `target' (i.e., a dot array of a particular cardinality) before the instruction, so that we could then refer back to the target's cardinality in the instruction using expressions like \emph{the same number of dots as the target}; \emph{fewer dots than the target}.
This presentation of a target before the main body of the trial shares some features with \citeauthor[Experiment 2]{Izard20081221}, although in that experiment participants were told the cardinality of the target (called an \emph{inducer} in that paper) whereas in our experiment we did not tell participants the cardinality of the prime array.
An item was thus a combination of a target dot array, an instruction that did not contain a number, and a set of dot arrays taking their cardinalities from the same triples used in Experiment 2.
Table \ref{Instructions for e4} spells out how the instructions were constrained not to mention a numeral and gives examples of targets.

\begin{table}[htp]
\caption{Instructions and targets by condition for Experiment 4}
\begin{center}
\begin{tabular}{llcrl}
\hline\noalign{\smallskip}
Item					    &Selection task					& Vagueness	& Target		&Choose a square with\ldots							\\
\noalign{\smallskip}\hline\noalign{\smallskip}
\multirow{4}{*}{06:15:24} 	&\multirow{2}{*}{Comparison} 	& Crisp		& 20			&\ldots fewer dots than the target					\\
\cline{3-5}
							&								& Vague		& 20			&\ldots far fewer dots than the target				\\
\cline{2-5}
							&\multirow{2}{*}{Matching}		& Crisp		& 6			&\ldots the same number of dots as the target		    \\
\cline{3-5}
							& 		 						& Vague		& 10 		&\ldots about the same number of dots as the target	    \\
\noalign{\smallskip}\hline
\end{tabular}
\end{center}
\label{Instructions for e4}
\end{table}%

\subsection{Hypotheses}% E 4
\noindent For Experiment 4, we hypothesised:

@@

\subsection{Results}% E 4
40 volunteers participated.
The results showed that vagueness was beneficial for comparison but detrimental for matching (the same as Experiment 3) even when no numbers were allowed in the instructions. 
Figure \ref{resultse4} shows the means by condition. Our findings were as follows:

@@

\begin{figure}[htbp]
\centering
\includegraphics[width=\textwidth]{figures/e4-rtplot-1.pdf}
\caption{Mean response times by condition for Experiment 4}
\label{resultse4}
\end{figure}

>>>>

\marginnote{E-Exp-3}

This experiment mirrors Experiment 2, but focussing on instructions that did not use a number. We manipulated vagueness and the selection task (comparison and matching). In order to implement the experiment without mentioning numbers in the instructions, we changed the sequence of each trial to include a `target' (i.e., a dot array of a particular cardinality) before the array, so that we could then refer to the target's cardinality in the instruction using expressions like \emph{the same number of dots as the target}; \emph{fewer dots than the target}. An example of this sequence is given in Figure \ref{Experiment3examplestimulus}. This presentation of a target before the main body of the trial shares some features with \citeauthor[Experiment 2]{Izard20081221}, although in that experiment participants were told the cardinality of the target (called an \emph{inducer} in that paper) whereas in our experiment we did not tell participants the cardinality of the prime array. An item was thus a combination of a target dot array, an instruction that did not contain a number, and a set of dot arrays taking their cardinalities from the same triples used in Experiments 1 and 2. Table \ref{instructionsE-exp-3} spells out how the instructions were constrained not to mention a numeral and gives examples of targets.

\begin{figure}[htbp]
\centering
\includegraphics[width=.75\textwidth]{figures/Ee3-flow.jpg}
\caption{Experiment 3 example stimulus}
\label{Experiment3examplestimulus}
\end{figure}

\begin{table}
\centering
\caption{Experiment 3 instructions arranged by condition. The instructions given in the table started with ``Choose a square with \ldots"}
\label{instructionsE-exp-3}
\resizebox{\textwidth}{!}{\begin{tabular}{ccccll}
\hline\noalign{\smallskip}
Item & Target & Quantity & Selection & Crisp & Vague \\ 
\noalign{\smallskip}\hline\noalign{\smallskip}
06:15:24 &   6 & Small & Matching 	& the same number of dots as the target & NA \\ 
06:15:24 &  10 & Small & Matching 	& NA & about the same number of dots as the target \\ 
06:15:24 &  10 & Large & Comparison & more dots than the target & far more dots than the target \\ 
06:15:24 &  20 & Small & Comparison & fewer dots than the target & far fewer dots than the target \\ 
06:15:24 &  20 & Large & Matching 	& NA & about the same number of dots as the target \\ 
06:15:24 &  24 & Large & Matching 	& the same number of dots as the target & NA \\ 
\noalign{\smallskip}\hline\noalign{\smallskip}
16:25:34 &  16 & Small & Matching 	& the same number of dots as the target & NA \\ 
16:25:34 &  20 & Small & Matching 	& NA & about the same number of dots as the target \\ 
16:25:34 &  20 & Large & Comparison & more dots than the target & far more dots than the target \\ 
16:25:34 &  30 & Small & Comparison & fewer dots than the target & far fewer dots than the target \\ 
16:25:34 &  30 & Large & Matching 	& NA & about the same number of dots as the target \\ 
16:25:34 &  34 & Large & Matching 	& the same number of dots as the target & NA \\ 
\noalign{\smallskip}\hline\noalign{\smallskip}
26:35:44 &  26 & Small & Matching 	& the same number of dots as the target & NA \\ 
26:35:44 &  30 & Small & Matching 	& NA & about the same number of dots as the target \\ 
26:35:44 &  30 & Large & Comparison & more dots than the target & far more dots than the target \\ 
26:35:44 &  40 & Small & Comparison & fewer dots than the target & far fewer dots than the target \\ 
26:35:44 &  40 & Large & Matching 	& NA & about the same number of dots as the target \\ 
26:35:44 &  44 & Large & Matching 	& the same number of dots as the target & NA \\ 
\noalign{\smallskip}\hline\noalign{\smallskip}
36:45:54 &  36 & Small & Matching 	& the same number of dots as the target & NA \\ 
36:45:54 &  40 & Small & Matching 	& NA & about the same number of dots as the target \\ 
36:45:54 &  40 & Large & Comparison & more dots than the target & far more dots than the target \\ 
36:45:54 &  50 & Small & Comparison & fewer dots than the target & far fewer dots than the target \\ 
36:45:54 &  50 & Large & Matching 	& NA & about the same number of dots as the target \\ 
36:45:54 &  54 & Large & Matching 	& the same number of dots as the target & NA \\ 
\noalign{\smallskip}\hline
\end{tabular}}
\end{table}

\subsection{Hypotheses}

For Experiment 3, we hypothesised:

@@

\subsection{Results}
40 volunteers participated. The results showed that vagueness was beneficial for comparison but detrimental for matching (the same as Experiment 2) even when no numbers were allowed in the instructions. Figure \ref{resultsE-exp-3} shows the means by condition. Our findings were as follows:

@@

\begin{figure}[htbp]
\centering
\includegraphics[width=\textwidth]{figures/Ee3-rtplot-1.pdf}
\caption{Mean response times by condition for Experiment 3}
\label{resultsE-exp-3}
\end{figure}
